---
title: "Pakistan_GLM"
author: "Volker Dickhardt"
date: "7 Januar 2020"
output: html_document
---
#1. Data preperation
#### Load packages:
```{r}
library("raster")
library("rgdal")
library("maptools")
library("geonames")
library("dismo")
```

#### Clear workspace
```{r}
rm(list=ls())
```

#### set working directory
```{r}
wd <- "E:/Uni/5. Semester/Pakistan_Projekt/bsc-sdm-2019-sharing/presentations/GLM_Volker"
setwd(wd)
dir()
```

#### sets a seed
```{r}
set.seed(0)
```

#### Read in the Shapefile "distribution_merged_Pakistan", These file shows the records of the individual species of butterflys in Pakistan.
```{r}
butterflys <- readOGR("E:/Uni/5. Semester/Pakistan_Projekt/bsc-sdm-2019/data/distribution/distribution_merged_Pakistan")
plot(butterflys)
```

#### This code extracts the coordinates of the records out of the shapefile and write them into a new dataframe called butterf
```{r}
head(butterflys)
but_cor<-coordinates(butterflys) # gives the coordinates
butterf<-as.data.frame(but_cor)
colnames(butterf) = c('lon','lat')

```

#### For this example we need the records of one species (Parnara bada)
```{r}
Parnara_bada <- butterf[111:145,] # extracts the coordinates of Parnara bada
```

#### Load Raster Data and stack the four Layers, plots the layers with the extent, Source: <http://chelsa-climate.org/downloads/>
```{r}
chel1 <- raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Chelsa/annual_global_prec/CHELSA_bio10_12.tif")
chel2 <- raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Chelsa/Annual_mean_Temperature/anual_mean_T.tif")
chel3 <- raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Chelsa/Precipitation of Warmest Quarter/CHELSA_bio10_18.tif")
chel4 <- raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Chelsa/warmest_quarter_mean_T_global/CHELSA_bio10_10.tif")

files<-c(chel1,chel2,chel3,chel4)
chelsa <- raster::stack(files) # stacks the Layers together

chelsa
nlayers(chelsa) # shows the number of layers in the raster stack
crs(chelsa) # shows the coordinate referent system of the layer
```

#### Load the Rasterlayer "Population of Pakistan"
```{r}
pop_pak <- raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/DivaGIS/pop_pak.tif")
pop_pak
crs(pop_pak) #shows the coordinate reference system of the Layer
```

#### Loads the three Worldclim-Layers, Source: <http://worldclim.org/version2>
```{r}
wc_tavg<-raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Worldclim/worldclim_temp_avg/tavg_mean.tif",pattern = ".tif", full.names = TRUE)
wc_tavg
wc_tmax<-raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Worldclim/worldclim_temp_max/tmax_mean.tif",pattern = ".tif", full.names = TRUE)
wc_tmax
wc_tmin<-raster("E:/Uni/5. Semester/Pakistan_Projekt/Daten/Raster/Worldclim/worldclim_temp_min/tmin_mean.tif",pattern = ".tif", full.names = TRUE)
wc_tmin #For this layers I calculated the mean in Q-GIS
```

#### Stack the worldclim layers together
```{r}
wc_layers <- c(wc_tavg,wc_tmax,wc_tmin)
wc<-stack(wc_layers)
```

#### Creates an extent with the extent of the Population Layer
```{r}
ext <- extent(c(60, 78, 23, 38))
```

#### With the comand "crop" you cut out a raster layer with a polygon, in this case with the extent.
```{r}
chelsa_ext<-crop(chelsa,ext)
pop_pak_ext<-crop(pop_pak,ext)
wc_ext<-crop(wc,ext)
```
#### Now we can stack the layers together
```{r}
b <- c(chelsa_ext,wc_ext)
predictors <- stack(b)
```

#### We can check the data by plotting a map with the first Layer of the predictors and all records
```{r}
data(wrld_simpl)
plot(predictors, 1)
plot(wrld_simpl, add=TRUE)
points(butterf, col='blue')
points(Parnara_bada, col="red", pch=18)
```

#### We need some background data for the absent data, This funktion creates random Points in the observation area
```{r}
backgr <- randomPoints(predictors, 500)
```

#### For a generalized linear model we need present and absent data. In the vektor "presvals" are the values of the predictor layers at the lokation of the records. In "absvals" are the values at the lokation of the randam points.
```{r}
presvals <- extract(predictors, Parnara_bada)
absvals <- extract(predictors, backgr)
```

#### We create a vektor (pb) witch writes for absent data 0 and for present data 1. Afterwards we build a dataframe with this vector and the present and absent data
```{r}
pb <- c(rep(1, nrow(presvals)), rep(0, nrow(absvals)))
sdmdata <- data.frame(cbind(pb, rbind(presvals, absvals)))
```

#### The command kfold gives every recort in the dataframe a number between 1 and k (in this case 5). We need this operation to get train and test data to test the model later on.
```{r}
group <- kfold(Parnara_bada, 5)
pres_train <- Parnara_bada[group != 1, ]
pres_test <- Parnara_bada[group == 1, ]
```

#### Now we do the same for the background data
```{r}
backg <- randomPoints(predictors, n=1000, ext=ext, extf = 1.05) # With extf the extent increases
colnames(backg) = c('lon','lat')
group <- kfold(backg, 5)
backg_train <- backg[group != 1, ]
backg_test <- backg[group == 1, ]
```

#### We create a matrix out of the present and background training data
```{r}
train <- rbind(pres_train, backg_train)
```

#### We build a dataframe with the training data and the predictor layers
```{r}
pb_train <- c(rep(1, nrow(pres_train)), rep(0, nrow(backg_train)))
envtrain <- extract(predictors, train)
envtrain <- data.frame( cbind(pa=pb_train, envtrain) )
head(envtrain) #shows the head of the dataframe
```

#### The same for the test data
```{r}
testpres <- data.frame( extract(predictors, pres_test) )
testbackg <- data.frame( extract(predictors, backg_test) )
```


#Build the model

#### We have everything to fit the generalized linear model
```{r}
m1 <- glm(pa ~ .,family = binomial(link = "logit"), data=envtrain)
m1
summary(m1)
```

#### Prediction with the model
```{r}
pg <- predict(predictors,m1, ext=ext)
plot(pg)
plot(wrld_simpl, add=TRUE)
points(Parnara_bada, col="red", pch=18)
```

#### Cross-validation of the model "m1"
```{r}
evaluate(testpres, testbackg, m1)
```


# References

#### Hijmans, R.H., Elith, J. (2019): Spatial Distribution Models. <https://rspatial.org/sdm/SDM.pdf> Date: 10.Jan.2020.
#### Hijmans, R.H. (2019): Point pattern analysis. <https://rspatial.org/raster/analysis/8-pointpat.html> Date: 08.Jan.2020.